Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-10 paper/paper.tex
minro

To the best of our knowledge, this paper presents the first Coq formalisation of
a recursion scheme, called the \emph{hylomorphism}, along with its algebraic
laws allowing for the mechanisation of all recognised (terminating) recursive
algorithms. The key contribution of this paper is that this formalisation is
fully axiom-free allowing for the extraction of safe, idiomatic OCaml code. We
exemplify the framework by formalising a series of algorithms based on different
recursive paradigms such as divide-and conquer, dynamic programming, and mutual
recursion and demonstrate that the extracted OCaml code for the programs
formalised in our framework is efficient, resembles code that a human programmer
would write, and contains no occurrences of \ocaml{Obj.magic}.  We also present
a machine-checked proof of the well-known short-cut fusion optimisation.
\end{abstract}

@@

\section{Formalising Recursion Schemes}
\label{sec:recursion-schemes}
\dcas{Do not cut below}
Recursion schemes provide an abstract way to consume and generate data. We now
proceed onto describing how to formalise hylomorphisms in Coq. We first
formalise algebras for container extensions (Section~\ref{sec:algebras}), then
we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together
these notions to formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:rec-coalgebras}).

>>>>

This paper presents the first Coq formalisation of
a recursion scheme, called the \emph{hylomorphism}, along with its algebraic
laws allowing for the mechanisation of all recognised (terminating) recursive
algorithms. The key contribution of this paper is that this formalisation is
fully axiom-free allowing for the extraction of safe, idiomatic OCaml code. We
exemplify the framework by formalising a series of algorithms based on different
recursive paradigms such as divide-and conquer, dynamic programming, and mutual
recursion and demonstrate that the extracted OCaml code for the programs
formalised in our framework is efficient, resembles code that a human programmer
would write, and contains no occurrences of \ocaml{Obj.magic}.  We also present
a machine-checked proof of the well-known short-cut fusion optimisation.
\end{abstract}

@@

\section{Formalising Recursion Schemes}
\label{sec:recursion-schemes}
Recursion schemes provide an abstract way to consume and generate data. We now
proceed onto describing how to formalise hylomorphisms in Coq. We first
formalise algebras for container extensions (Section~\ref{sec:algebras}), then
we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together
these notions to formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:rec-coalgebras}).
